Implications of uniformly distributed, empirically informed priors for phylogeographical model selection: A reply to Hickerson et al.

### 2014-08-05 manuscripts/reply/reply-abstract.tex
More tweaking of reply ms.

Establishing that a set of population splitting events occurred at the same
time can be a potentially persuasive argument that the populations were
affected by the same geographic event.
Recently, \citet{Oaks2012} assessed the ability of an approximate-Bayesian
model choice method (\msb) to estimate such patterns of divergences across
taxa, to which \citet{Hickerson2013} responded.
Both papers agree that the primary inference enabled by the method is very
sensitive to prior assumptions and often erroneously supports shared
divergences across taxa.
However, the papers differ in the best explanation and solution for this
problem.
\citet{Hickerson2013} suggested the lack of robustness in \msb analyses is
caused by numerical approximation error due to the inefficiency of the method's
rejection algorithm under broad uniform priors on divergence times
(Hypothesis~\ref{hypError}).
To overcome this, they proposed a model-averaging approach that uses narrow,
empirically informed uniform priors.
Here we demonstrate the approach of \citet{Hickerson2013} will often
erroneously support shared divergence events and is dangerous in the sense that
the empirically-derived uniform priors often exclude from consideration the
true values of the divergence-time parameters.
\citet{Oaks2012} suggested the method's behavior was due to the sensitivity of
marginal likelihoods to the uniformly distributed priors on divergence times
(Hypothesis~\ref{hypML}).
To mitigate this sensitivity, they suggested alternative probability
distributions to accommodate prior uncertainty about divergence times.
As predicted by fundamental principles of Bayesian model choice, more flexible
distributions can accommodate prior uncertainty about parameters without
placing excessive prior weight in vast regions of parameter space with low
likelihood;
this increases the method's robustness and power to detect temporal variation
in divergences \citep{Oaks2014dpp}.
We also show that the tendency of \msb analyses to support models of shared
divergences is primarily due to Hypothesis~\ref{hypML};
Hypothesis~\ref{hypError} is not a tenable explanation for the bias.
All of these papers have highlighted the importance of choosing prior
assumptions about parameters in Bayesian model selection, because they can
strongly influence the posterior probabilities of the models we seek to
estimate.
No amount of computation can rescue our inference if we choose priors on
parameters such that the exact posterior supports the wrong model of
evolutionary history.

>>>>

Establishing that a set of population splitting events occurred at the same
time can be a potentially persuasive argument that the populations were
affected by the same geographic event.
Recently, \citet{Oaks2012} assessed the ability of an approximate-Bayesian
model choice method (\msb) to estimate such patterns of divergences across
taxa, to which \citet{Hickerson2013} responded.
Both papers agree that the primary inference enabled by the method is very
sensitive to prior assumptions and often erroneously supports shared
divergences across taxa when prior uncertainty about divergence times
is represented by a uniform distribution.
However, the papers differ in the best explanation and solution for this
problem.
\citet{Hickerson2013} suggested the lack of robustness in \msb analyses is
caused by numerical approximation error due to the inefficiency of the method's
rejection algorithm under broad uniform priors on divergence times
(Hypothesis~\ref{hypError}).
To overcome this, they proposed a model-averaging approach that uses narrow,
empirically informed uniform priors.
Here we demonstrate the approach of \citet{Hickerson2013} will often
erroneously support shared divergence events and is dangerous in the sense that
the empirically-derived uniform priors often exclude from consideration the
true values of the divergence-time parameters.
\citet{Oaks2012} suggested the method's behavior was due to the sensitivity of
marginal likelihoods to the uniformly distributed priors on divergence times
(Hypothesis~\ref{hypML}).
To mitigate this sensitivity, they suggested alternative probability
distributions to accommodate prior uncertainty about divergence times.
As predicted by fundamental principles of Bayesian model choice, more flexible
distributions can accommodate prior uncertainty about parameters without
placing excessive prior weight in vast regions of parameter space with low
likelihood;
this increases the method's robustness and power to detect temporal variation
in divergences \citep{Oaks2014dpp}.
We also show that the tendency of \msb analyses to support models of shared
divergences is primarily due to Hypothesis~\ref{hypML}, whereas
Hypothesis~\ref{hypError} is not a tenable explanation for the bias.
All of these papers have highlighted the importance of choosing prior
assumptions about parameters in Bayesian model selection, because they can
strongly influence the posterior probabilities of the models we seek to
estimate.
No amount of computation can rescue our inference if we choose priors on
parameters such that the exact posterior supports the wrong model of
evolutionary history.

### 2014-08-05 manuscripts/reply/reply-main.tex
More tweaking of reply ms.

Recently, \citet{Oaks2012} used this model-choice framework to study 22 pairs
of vertebrate lineages distributed across the Philippines; they also studied
the behavior of the \msb approach using computer simulations.
They found the method is very sensitive to prior assumptions and often
supports shared divergences across taxa that diverged randomly over broad time
periods, to which \citet{Hickerson2013} responded.
\citet{Oaks2012} and \citet{Hickerson2013} agree on the fundamental
methodological point about the model selection performed in \msb:
\begin{itemize}
   \item The use of vague priors on the divergence-time parameters can lead to
       spurious support for models with few divergence events shared across
       taxa. Thus, the primary inference enabled by the approach is very
       sensitive to the priors on divergence times.
\end{itemize}
However, the two papers suggest alternative mechanisms by which the priors on
divergence times cause this behavior:
\begin{enumerate}[label=\textsl{\textbf{Hypothesis \arabic*)}},ref=\arabic*,align=left]
    \item \textsl{\textbf{Numerical approximation error}} \citep{Hickerson2013}:
        Under broad uniform priors, the rejection algorithm implemented in \msb
        is unable to adequately sample the space of the models within
        reasonable computational time, which leads to bias toward models with
        fewer divergence-time parameters because they are better sampled.
        \label{hypError}
    \item \textsl{\textbf{Strongly weighted marginal likelihoods}}
        \citep{Oaks2012}:
        The uniform priors on divergence times lead to very small marginal
        likelihoods (and thus posterior probabilities) of models with many
        divergence-time parameters.  The likelihood of these models is
        ``averaged'' over a much greater parameter space in which there is a
        large amount of prior weight and small probability of producing the
        data.  \citep{Jeffreys1939,Lindley1957}. \label{hypML}
\end{enumerate}
In Hypothesis~\ref{hypError}, the problem is numerical approximation error due to
insufficient computation;
given data from randomly diverged taxa, the exact posterior supports a model
with many divergence-time parameters, but we are unable to accurately
approximate this posterior.
In Hypothesis~\ref{hypML}, the problem is more fundamental; given data from
randomly diverged taxa, the exact posterior supports a model with a small
number of simultaneous divergences.
I.e., given the assumption that divergence times are broadly and uniformly
distributed, the exact posterior from Bayes' rule leads us to the wrong
conclusion about evolutionary history.
Such posterior support for simultaneous divergence, even if ``correct'' from
the perspective of Bayesian model choice, does not provide the biogeographical
insights that a researcher who employs \msb seeks to gain.

While these phenomena are not mutually exclusive, it is important to
distinguish between them in order to determine how to improve our ability to
estimate shared divergence histories.
If Hypothesis~\ref{hypError} is correct, then the model is sound and we need to
increase our computational effort or improve our Monte Carlo integration
procedures.
For example, Markov chain or sequential Monte Carlo algorithms might sample the
posterior more efficiently than the simple Monte Carlo rejection sampler
implemented in \msb.
Rather than alter the sampling algorithm, \citet{Hickerson2013} tried using
narrow, empirically informed uniform priors in the hope that with less
parameter space to sample, the rejection algorithm would produce better
estimates of the posterior.
Here, we discuss theoretical considerations for using empirically informed
priors for Bayesian model choice and evaluate the approach of
\citet{Hickerson2013} as a potential solution to the biases of \msb revealed by
\citet{Oaks2012}.
In their analyses, \citet{Hickerson2013} made an error by mixing different
units of time, which invalidates the results presented in their response (see
Supporting Information for details).
We correct this error, but still find their approach will often support
(1) clustered divergence models when divergences are random, and
(2) models that exclude from consideration the true values of the
parameters.

@@

We also explore the distinct predictions made by Hypotheses 1 and 2.
We show the behavior of \msb matches the predictions of Hypothesis~\ref{hypML},
but not Hypothesis~\ref{hypError}.
This strongly suggests the method tends to support models of shared divergences
not because of insufficient computation, but rather due to the larger marginal
likelihoods of these models under the prior assumption of uniformly distributed
divergence times.

@@

However, \citet{Hickerson2013} conclude that the bias is caused by numerical
approximation error due to insufficient computation
(Hypothesis~\ref{hypError}).
They argue the widest of the three priors on divergence times used by
\citet{Oaks2012} would infrequently produce random samples with many
independent population divergence times as recent as the estimated gene
divergence times presented in \citet{Oaks2012}.
% They also also suggest that Oaks et al.'s \citeyear{Oaks2012} estimated gene
% divergence times should be used to set the upper bound of the uniform prior on
% divergence times for the \msb analysis of the same sequence data.
% Given the gene-tree estimates were intended to provide only a rough comparison
% of the gene divergence times across the 22 taxa \citep{Oaks2012}, and our
% findings presented above regarding the behavior of \msb when empirically
% informed uniform priors are employed, such an approach should not be used.
However, this sampling-probability argument is based on some questionable
assumptions.
Oaks et al.'s (\citeyear{Oaks2012}) gene-tree estimates were intended to
provide only a rough comparison of the gene divergence times across the 22 taxa
and assumed an arbitrary strict per-site rate of $2\e{-8}$ mutations per
generation for all taxa.
Furthermore, because the branch-length units of the gene trees are in millions
of years whereas the divergence-time prior of \msb is in generations,
\citet{Hickerson2013} make the implicit assumption that all 22 Philippine taxa
have a generation time of one year.
More importantly, even if we assume (1) the arbitrary strict clock is correct,
(2) gene divergence times were estimated without error, and (3) all 22 taxa
have one-year generation times, Hickerson et al.'s (\citeyear{Hickerson2013})
argument actually demonstrates that the models used by \citet{Oaks2012} with
narrower priors on divergence times are densely populated with samples with
large numbers of divergence parameters with values younger than the estimated
gene divergence estimates.
Thus, if \citet{Hickerson2013} are correct, analyses under these narrow priors
should be much less biased toward clustered models of divergence.
However, the magnitude of the bias is very similar across all three priors
explored by \citet{Oaks2012}.
\citet{Hickerson2013} point out a case where the narrowest prior performs
slightly better (panel L of Figures S32, S37, and S38 of \citet{Oaks2012}).
However, it is important to note that these results suffered from a bug in
\msb, and after \citet{Oaks2012} corrected the bug, there are many cases where
the narrowest prior performs slightly worse (see panels D--J of Figures 3 and
S12).

To disentangle whether Hypothesis 1 or 2 is the primary cause the method's
erroneous support for simple models, we must look at the different predictions
made by these two phenomena.
For example, numerical error due to insufficient prior sampling
(Hypothesis~\ref{hypError}) should create large variance among posterior
estimates and cause analyses to be highly sensitive to the number of samples
drawn from the prior.
Furthermore, if insufficient prior sampling is \emph{biasing} estimates toward
models with less parameter space we expect to see support for these models
decrease as sampling from the prior increases.
\citet{Oaks2012} did not see such sensitivity when they compared prior sample
sizes of $2\e6$, $5\e6$, and $10^7$.

@@

This is further demonstrated by the results presented herein that show the
model-averaging approach of \citet{Hickerson2013} prefers models with narrower
\divt{} priors (Table~\ref{tabModelChoiceEmpirical} and
Figs.~\labelcref{figExclusionSimTau,figExclusionSimProb,figPowerExclusion4})
and fewer \divt{} parameters (Figure~\ref{figPower4}).
For these model-averaging analyses, insufficient prior sampling
(Hypothesis~\ref{hypError}) is an unlikely explanation for the erroneous support
for models with less parameter space, because (1) all of the prior models share
the same dimensionality, and (2) the same number of random samples were drawn
from each of the prior models.
These results are predicted by Hypothesis~\ref{hypML}, because the marginal
likelihoods will be higher for models with narrower priors on divergence times
and fewer divergence-time dimensions (these models integrate over less space
with large prior weight and small likelihood).

>>>>

Recently, \citet{Oaks2012} used this model-choice framework to study 22 pairs
of vertebrate lineages distributed across the Philippines; they also studied
the behavior of the \msb approach using computer simulations.
They found the method is very sensitive to prior assumptions and often
supports shared divergences across taxa that diverged randomly over broad time
periods, to which \citet{Hickerson2013} responded.
\citet{Oaks2012} and \citet{Hickerson2013} agree on the fundamental
methodological point about the model selection performed in \msb:
\begin{itemize}
   \item Representing prior uncertainty about divergence-time parameters with a
       uniform distribution can lead to spurious support for models with few
       divergence events shared across taxa. Thus, the primary inference
       enabled by the approach is very sensitive to the priors on divergence
       times.
\end{itemize}
However, the two papers suggest alternative mechanisms by which the priors on
divergence times cause this behavior:
\begin{enumerate}[label=\textsl{\textbf{Hypothesis \arabic*)}},ref=\arabic*,align=left]
    \item \textsl{\textbf{Numerical approximation error}} \citep{Hickerson2013}:
        Under broad uniform priors, the rejection algorithm implemented in \msb
        is unable to adequately sample the space of the models within
        reasonable computational time, which leads to bias toward models with
        fewer divergence-time parameters because they are better sampled.
        \label{hypError}
    \item \textsl{\textbf{Strongly weighted marginal likelihoods}}
        \citep{Oaks2012}:
        The uniform priors on divergence times lead to very small marginal
        likelihoods (and thus posterior probabilities) of models with many
        divergence-time parameters.  The likelihood of these models is
        ``averaged'' over a much greater parameter space in which there is a
        large amount of prior weight and small probability of producing the
        data.  \citep{Jeffreys1939,Lindley1957}. \label{hypML}
\end{enumerate}
In Hypothesis~\ref{hypError}, the problem is numerical approximation error due
to insufficient computation.
In this scenario, given data from taxa that diverged randomly through time, the
exact posterior supports a model with many divergence-time parameters, but we
are unable to accurately approximate this posterior.
In Hypothesis~\ref{hypML}, the problem is more fundamental; given data from
taxa that diverged randomly through time, the exact posterior supports a model
with simultaneous divergences across taxa.
I.e., when accommodating prior uncertainty about divergence times with a
uniform distribution, the exact posterior from Bayes' rule leads us to the
wrong conclusion about evolutionary history.
Such posterior support for simultaneous divergence, even if ``correct'' from
the perspective of Bayesian model choice, does not provide the biogeographical
insights that a researcher who employs \msb seeks to gain.

While these phenomena are not mutually exclusive, it is important to
distinguish between them in order to determine how to improve our ability to
estimate shared divergence histories.
If Hypothesis~\ref{hypError} is correct, then the model is sound and we need to
increase our computational effort or improve our Monte Carlo integration
procedures.
For example, Markov chain or sequential Monte Carlo algorithms might sample the
posterior more efficiently than the simple Monte Carlo rejection sampler
implemented in \msb.
Rather than alter the sampling algorithm, \citet{Hickerson2013} tried using
narrow, empirically informed uniform priors in the hope that with less
parameter space to sample, the rejection algorithm would produce better
estimates of the posterior.
Here, we discuss theoretical considerations for using empirically informed
priors for Bayesian model choice and evaluate the approach of
\citet{Hickerson2013} as a potential solution to the biases of \msb.
In their analyses, \citet{Hickerson2013} made an error by mixing different
units of time, which invalidates the results presented in their response (see
Supporting Information for details).
We correct this error, but still find their approach will often support
(1) clustered divergence models when divergences are random, and
(2) models that exclude from consideration the true values of the
parameters.

@@

We also explore the distinct predictions made by Hypotheses 1 and 2.
We show the behavior of \msb matches the predictions of Hypothesis~\ref{hypML},
but not Hypothesis~\ref{hypError}.
This strongly suggests that the method tends to support models of shared
divergences not because of insufficient computation, but rather due to the
larger marginal likelihoods of these models under the prior assumption of
uniformly distributed divergence times.

@@

However, \citet{Hickerson2013} conclude that the bias is caused by numerical
approximation error due to insufficient computation
(Hypothesis~\ref{hypError}).
They argue the widest of the three priors on divergence times used by
\citet{Oaks2012} would infrequently produce random samples of parameter values
with many independent population divergence times as recent as the estimated
gene divergence times presented in \citet{Oaks2012}.
% They also also suggest that Oaks et al.'s \citeyear{Oaks2012} estimated gene
% divergence times should be used to set the upper bound of the uniform prior on
% divergence times for the \msb analysis of the same sequence data.
% Given the gene-tree estimates were intended to provide only a rough comparison
% of the gene divergence times across the 22 taxa \citep{Oaks2012}, and our
% findings presented above regarding the behavior of \msb when empirically
% informed uniform priors are employed, such an approach should not be used.
However, this sampling-probability argument is based on some questionable
assumptions.
Oaks et al.'s (\citeyear{Oaks2012}) gene-tree estimates were intended to
provide only a rough comparison of the gene divergence times across the 22 taxa
and assumed an arbitrary strict per-site rate of $2\e{-8}$ mutations per
generation for all taxa.
Furthermore, because the branch-length units of the gene trees are in millions
of years whereas the divergence-time prior of \msb is in generations,
\citet{Hickerson2013} make the implicit assumption that all 22 Philippine taxa
have a generation time of one year.
More importantly, even if we assume (1) the arbitrary strict clock is correct,
(2) gene divergence times were estimated without error, and (3) all 22 taxa
have one-year generation times, Hickerson et al.'s (\citeyear{Hickerson2013})
argument actually demonstrates that the models used by \citet{Oaks2012} with
narrower priors on divergence times are densely populated with samples with
large numbers of divergence parameters with values younger than the estimated
gene divergence estimates.
Thus, if \citet{Hickerson2013} are correct, analyses under these narrow priors
should be much less biased toward clustered models of divergence.
However, the magnitude of the bias is very similar across all three priors
explored by \citet{Oaks2012}.
\citet{Hickerson2013} point out a case where the narrowest prior performs
slightly better (panel L of Figures S32, S37, and S38 of \citet{Oaks2012}).
However, it is important to note that these results suffered from a bug in
\msb, and after \citet{Oaks2012} corrected the bug, there are many cases where
the narrowest prior performs slightly worse (see panels D--J of Figures 3 and
S12).

To disentangle whether Hypothesis 1 or 2 is the primary cause of the method's
erroneous support for simple models, we must look at the different predictions
made by these two phenomena.
For example, numerical error due to insufficient prior sampling
(Hypothesis~\ref{hypError}) should create large variance among posterior
estimates and cause analyses to be highly sensitive to the number of samples
drawn from the prior.
Furthermore, if insufficient prior sampling is \emph{biasing} estimates toward
models with less parameter space we expect to see support for these models
decrease as sampling from the prior increases.
\citet{Oaks2012} did not see such sensitivity when they compared prior sample
sizes of $2\e6$, $5\e6$, and $10^7$.

@@

This is further demonstrated by the results presented herein that show the
model-averaging approach of \citet{Hickerson2013} prefers models with narrower
\divt{} priors (Table~\ref{tabModelChoiceEmpirical} and
Figs.~\labelcref{figExclusionSimTau,figExclusionSimProb,figPowerExclusion4})
and fewer \divt{} parameters (Figure~\ref{figPower4}).
For these model-averaging analyses, insufficient prior sampling
(Hypothesis~\ref{hypError}) is an untenable explanation for the erroneous support
for models with less parameter space, because (1) all of the prior models share
the same dimensionality, and (2) the same number of random samples were drawn
from each of the prior models.
However, these results are predicted by Hypothesis~\ref{hypML}, because the
marginal likelihoods will be higher for models with narrower priors on
divergence times and fewer divergence-time dimensions (these models integrate
over less space with large prior weight and small likelihood).
